Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-12 content/contribution/continuous-datalog.tex
Fixed up to the end of defn of TA-Datalog

\[\alpha_0 :\!-\ \alpha_1 , \ldots , \alpha_n\]
where $\alpha_i$ are atomic formulae.

@@

We define the concept of entailment in Datalog as follows. A fact $f$
is entailed by a program $\Pi$ iff each interpretation satisfying
$\Pi$ (satisfying every clause in $\Pi$) also satisfies $f$, written
$\Pi \models f$. The entailment of $\Pi$ is therefore the set of all
facts that logically follow from $\Pi$:

\[cons(\Pi) = \{ f \in \hat{H} | \Pi \models f \}\]

%For convenience, we will write $\Pi \models E$ to indicate that $E =
%cons(\Pi)$ is the set of all facts entailed by $\Pi$.

@@

We define a {\em justification} of an entailment of a Datalog program
as a minimal subset of the program that is sufficient for the
entailment to hold. More precisely, for a Datalog program
$\Pi = \langle R, F \rangle$ and an entailment $E$ where
$\langle R, F \rangle \models E$, the program $\langle J_R, J_F \rangle$
with $J_R \subseteq R$ and $J_F \subseteq F$ is a justification for $E$
iff $\langle J_R, J_F \rangle \models E$ and, for all $J'_R \subset J_R$ or $J'_F \subset J_F$, $\langle J'_R, J'_F \rangle \not\models E$.

We can simplify this notion of justification by making the assumption
that the justification can be defined in terms of a minimal subset of
the ground facts of the program; the rules of the program are
therefore implicitly considered to be part of every justification. We
refer to this simplified justification as a {\em fact justification},
and define it as follows: given a Datalog program $\Pi = \langle R, F
\rangle$ and an entailment $E$ where $\langle R, F \rangle \models E$,
the set of facts $J \subseteq F$ is a fact justification for $E$ iff
$\langle R, J \rangle \models E$ and, for all $J' \subset J$, $\langle R, J' \rangle
\not\models E$.
\end{definition}

@@

Let $\mathbb{T}$ be a discrete time domain with a total order under
$\leqslant$; an element of $\mathbb{T}$ is an {\em instant}.
$\mathbb{T}$ has a lower bound $0$, such that for all $i \in
\mathbb{T}$, $0 \leqslant i$. For convenience, we write $i+n$ where $i
\in \mathbb{T}$ to indicate the $n$th successor of instant $i$,
similarly $i-n$.

%It should be noted that time is a continuous concept, with infinitely
%many divisions possible between any two discrete instants.  However,
%the resolution with which such divisions can be recognised will, in
%practise, have a lower bound set by hardware limitations, or may be
%defined as part of some specific implementation.  In these cases,
%anything that occurs between discrete instants is considered to have
%occurred at the next recognised instant.
\end{definition}

@@

We represent the relationship between a fact (without temporal
annotation) and a fact instance (with temporal annotation) with the
function ${strip : \hat{H} \times \mathbb{T} \rightarrow \hat{H}}$
which maps a fact instance to its underlying fact:

\[ strip(\langle f, t\rangle) = f \]

For convenience, we also write $strip(F^T) = F$, where $F^T$ is a set
of fact instances and $F = \{ strip(f) | f \in F^T \}$.

@@

@@

A time-annotated Datalog program $\Pi^T = \langle R, F^T\rangle$ is a
tuple which comprises a finite set of rules $R$ and a finite set of
fact instances $F^T \subseteq \hat{H} \times \mathbb{T}$. Fact
instances are as defined in \refdef{continuous datalog: fact
  instance}, and rules are of the same form as in conventional
Datalog programs.

When considering a time-annotated Datalog program, in which there may
be many distinct instances of a given fact, temporal information
regarding when instances of that fact became true would be lost if
such a program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts at all
instants). As before, a fact $f$ is entailed by a program $\Pi^T$ iff
each interpretation satisfying every clause in $\Pi^T$ also satisfies
$f$. The entailment of $\Pi^T$ is therefore:

\[ cons^T(\Pi^T) = \{ f \in \hat{H}^T | \Pi^T \models f \} \]

It follows that applying the ${strip()}$ function to all the fact
instances in both the set of fact instances $F^T$ in some ${\Pi^T =
  \langle R, F^T \rangle}$ and the set of entailed fact instances
$cons^T(\Pi^T)$ maintains the relationship of entailment; if a fact
instance $f$ is entailed by $\langle R, F^T\rangle$, then $strip(f)$
will be entailed by $\langle R, strip(F^T) \rangle$.
\end{definition}

@@

The justification of entailments (as in \refdef{continuous datalog:
  justification}) is equally applicable to time-annotated Datalog
programs, with facts in the entailment and the justification for that
entailment being replaced with fact instances.

Furthermore, each justification for $f$ in some $F^T$
entails a unique instance ${f | t}$, such that many instances of $f$
may occur in the set $E^T$ of fact instances entailed by ${\Pi^T =
  \langle R, W^{S,x}_i \rangle}$, each with a single justification.

As such, entailment justification for fact instances may be defined
formally by \refdef{continuous datalog: instance justification}.

A justification for some entailment ${f|t}$ of some time-annotated Datalog program
	${\Pi^T = \langle R , F^T \rangle \vdash f|t}$ is the single minimal set ${J^T \subseteq F^T}$
	such that ${\Pi^{T\prime} = \langle R , J^T \rangle \vdash f|t}$.
	The mapping of the fact instance ${f | t}$ to the minimal subset $J^T$
	within ${\Pi^T = \langle R, F^T \rangle}$ that justifies it is expressed
	as the function ${just : \mathbb{\Pi^T} , \mathbb{F^T} \rightarrow \{ \mathbb{F^T} \}}$.
\end{definition}

This allows us to formally specify that the time $t$ at which an entailed fact
instance ${f | t}$ becomes true is the last time at which any of its
justifying instances become true:
\[ t = max(\{ t^\prime : f^\prime | t^\prime \in just(\Pi^T, f | t) \}) \]
As such, the set of timestamps $T$ for some set $E^T$ of instances entailed by some
program ${\langle R, F^T \rangle}$ is a subset of the timestamps for the instances in $F^T$.

\begin{definition}[Stream of Facts]
\labeldef{continuous datalog: stream}
A stream of facts $S$ is a \emph{sequence} of fact instances ordered
by their timestamps; in conjunction with the distinction between
multiple instances of a fact by their timestamp, this definition justifies
Axiom~\ref{axiom:continuous datalog: window disjointness}.
\end{definition}

@@

A sliding window of range ${x \in \mathbb{N}}$ over any stream $S$ at
any given instant ${i \in \mathbb{N}}$ is the \emph{set} of fact
instances that occur in $S$ during the interval ${(i-x,i]}$, written
${W^{S,x}_{i}}$.

\[ W^{s,x}_i = \{ \langle f, t \rangle | \langle f, t \rangle \in S \land (i-x) < t \leqslant i \}\]

@@

\begin{axiom}\label{axiom:continuous datalog: window range leq 0}
A window of range $0$ over a stream $S$ at an instant 
${i \in \mathbb{N}}$ is always empty.
\begin{equation*}
W^{S,0}_{i} = \varnothing
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window composition}
A window of range ${x \ge 0}$ over a stream $S$ at an instant 
${i \in \mathbb{N}}$ is equal to the union of the $x$ consecutive
windows of range $1$ over the stream $S$ preceding and including that at
instant $i$.
\begin{equation*}
W^{S,x}_{i} = \mathop{\cup}_{j=0}^{x-1} W^{S,1}_{i-j}
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window disjointness}
A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an
instant ${i \in \mathbb{N}}$ is disjoint from a second window of 
range ${y \in \mathbb{N}}$ over $S$ at any instant before $i - x$.
\begin{equation*}
W^{S,x}_{i} \cap W^{S,y}_{i-z} = \varnothing
\end{equation*}
where $z \geq x$.
\end{axiom}

In \refdef{continuous datalog: datalog program}, we took a Datalog
program to consist of a set of rules and a set of facts.  In order to
extend this definition of Datalog programs to incorporate fact
instances, as in \refdef{continuous datalog: fact instance}, we define
a time-annotated Datalog program as follows:

>>>>

\[ \alpha_0 \Leftarrow \alpha_1 \land \ldots \land \alpha_n \]

\noindent where $\alpha_i$ are atomic formulae. Note that we have excluded any
universally quantified variables for clarity.

@@

We define the concept of entailment in Datalog as follows. A fact $e$
is entailed by a program $\Pi$ iff each interpretation satisfying
$\Pi$ (satisfying every clause in $\Pi$) also satisfies $e$, written
$\Pi \models e$. For a a program $\Pi$ and a set of entailed facts
$E$, $\Pi \models E$ iff $\Pi \models e$ for every $e \in E$. For
convenience, we refer to the set of all facts that are entailed by a
program (the intersection of all the Herbrand models of the program)
as the {\em entailment of a program}, written $cons(\Pi)$ and defined
as:
\begin{align*}
  cons(\Pi) &= \{ e \in \hat{H} \mid \Pi \models e \} \\
  &= \bigcap_\mathcal{I} \models_\mathcal{I} \Pi
\end{align*}
\end{definition}

We define a {\em justification} of a set of facts that are entailed by
a Datalog program as a minimal subset of the program that is
sufficient for the entailment to hold. More precisely, for a Datalog
program $\Pi = \langle R, F \rangle$ and a set of entailed facts $E$
where $\langle R, F \rangle \models E$, the program $\langle J_R, J_F
\rangle$ with $J_R \subseteq R$ and $J_F \subseteq F$ is a
justification for $E$ iff $\langle J_R, J_F \rangle \models E$ and,
for all $J'_R \subset J_R$ or $J'_F \subset J_F$, $\langle J'_R, J'_F
\rangle \not\models E$.

We simplify this notion of justification by making the assumption that
a justification can be defined in terms of a minimal subset of the
ground facts of the program; the rules of the program are therefore
implicitly considered to be part of every justification. Any
subsequent references to {\em justification} are to this simplified
sense, which is defined as follows: given a Datalog program $\Pi =
\langle R, F \rangle$ and a set of entailed facts $E$ where $\langle
R, F \rangle \models E$, the set of facts $J \subseteq F$ is a
justification for $E$ iff $\langle R, J \rangle \models E$ and, for
all $J' \subset J$, $\langle R, J' \rangle \not\models E$. Note that a
given set of entailed facts may not have a unique justification.

@@

@@

Let $\mathbb{T}$ be a discrete time domain with a total order under
$\leqslant$; an element of $\mathbb{T}$ is an {\em instant}.
$\mathbb{T}$ has a lower bound $0$, such that for all $i \in
\mathbb{T}$, $0 \leqslant i$. For convenience, we write $i+n$ where $i
\in \mathbb{T}$ to indicate the $n$th successor of instant $i$,
and $i-n$ to indicate the $n$th predecessor of instant $i$.

@@

We represent the relationship between a fact (without temporal
annotation) and a fact instance (with temporal annotation) with the
function $strip()$ which maps a fact instance to its underlying fact:
\[ strip(\langle f, t\rangle) = f \]

For convenience, we also write $strip(F^T) = F$, where $F^T$ is a set
of fact instances and $F = \{ strip(f) \mid f \in F^T \}$.

@@

In \refdef{continuous datalog: datalog program}, we took a Datalog
program to consist of a set of rules and a set of facts.  In order to
extend this definition of Datalog programs to incorporate fact
instances, we define a Time-Annotated Datalog program as follows:

@@

A Time-Annotated Datalog program (henceforth abbreviated TA-Datalog)
$\Pi^T = \langle R, F^T\rangle$ is a tuple which comprises a finite
set of rules $R$ and a finite set of fact instances $F^T \subseteq
\hat{H} \times \mathbb{T}$. Fact instances are as defined in
\refdef{continuous datalog: fact instance}, and rules are of the same
form as in conventional Datalog programs.

When considering a TA-Datalog program, in which there may
be many distinct instances of a given fact, temporal information
regarding when instances of that fact became true would be lost if
such a program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts at all
instants). As before, a fact instance $e$ is entailed by a program
$\Pi^T$ iff each interpretation satisfying every clause in $\Pi^T$
also satisfies $e$. The entailment of the program $\Pi^T$ is therefore:
\[ cons^T(\Pi^T) = \{ e \in \hat{H}^T \mid \Pi^T \models e \} \]

Informally, the set of instants appearing in some set $E^T$ of fact
instances entailed by some program ${\langle R, F^T \rangle}$ will be
a subset of the instances appearing the fact instances in $F^T$.  In
addition, the instant $t$ in a fact instance $\langle e, t \rangle$
that is entailed by a TA-Datalog program $\Pi^T = \langle R^T, F^T
\rangle$ will be bounded from above by the maximum instant in any fact
instance in $F^T$. A TA-Datalog program may entail fact instances that
are only distinguished by their instants (and are therefore considered
distinct from each other, as per \refdef{continuous datalog: fact
  instance}).

Consider the following example TA-Datalog program:
\begin{align*}
  \Pi^T = \langle {} & \{ a \Leftarrow b \land c \}, \\
  & \{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle  \} \rangle
\end{align*}

The entailment of the program $\Pi^T$ will be:

\[cons^T(\Pi^T) = \{\langle c, 1 \rangle, \langle c, 2 \rangle\}\]

It follows that applying the ${strip()}$ function to all the fact
instances in both the set of fact instances $F^T$ in some ${\Pi^T =
  \langle R, F^T \rangle}$ and the entailment of the program $\Pi^T$
(the set of fact instances $cons^T(\Pi^T)$) maintains the relationship
of entailment; if $\langle R, F^T\rangle \models \langle e, t
\rangle$, then $\langle R, strip(F^T) \rangle \models strip(\langle e,
t \rangle)$.

This relationship is shown in \reffig{TD-D}.

\begin{figure}[htb]
  \[
  \begin{CD}
\Pi^T = \langle R, F^T \rangle @>\models>> cons^T(\Pi^T) \\
@V{strip}VV @V{strip}VV \\
\Pi = \langle R, F \rangle @>\models>> cons(\Pi)
  \end{CD}
  \]
\caption{Relationship between Datalog and TA-Datalog entailments}
\labelfig{TD-D}
\end{figure}

Reconsidering the example TA-Datalog program above, we may convert it
into the following equivalent conventional Datalog program:
\[
  \Pi = \langle  \{ a \Leftarrow b \land c \}, \{ a, b \} \rangle
\]
Trivially, the entailment of this program $\Pi$ will therefore be:
\[cons(\Pi) = \{ c \}\]
\end{definition}

@@

The justification of entailments is equally applicable to TA-Datalog
programs, with facts in the entailment and the justification for that
entailment being replaced with fact instances.  Following the
simplified notion of justification in \refdef{continuous datalog:
  justification}, we define the justification of a set of fact
instances that are entailed by a TA-Datalog as a minimal subset of
the ground fact instances of the program that are sufficient for
the entailment to hold.

More formally, for a TA-Datalog program $\Pi^T = \langle R, F^T \rangle$
and a set of entailed fact instances $E^T$ where $\langle R, F^T \rangle \models E^T$, the set of fact instances $J^T \subseteq F^T$ is a justification for
$E^T$ iff $\langle R, J^T \rangle \models E^T$ and, for all $J^{T\prime} \subset J^T$, $\langle R, J^{T\prime} \not\models E^T$.

For an entailed fact instance $\langle e, t\rangle$ and a
justification of that fact instance $J^T$, the instant $t$ at which
$\langle e, t \rangle$ becomes true is take to be the latest instant
at which any of the fact instances in the justification became true:
\[ t = max(\{ t' \mid \langle j', t' \rangle \in J^T \}) \]

Note that, although an entailed fact instance $\langle e, t\rangle$
may have multiple justifications, all the justifications for $\langle
e, t \rangle$ will share the same maximum instant; a justification
with a different maximum instant $t' \neq t$ would entail a different
fact instance $\langle e, t' \rangle$. Considering again the example
TA-Datalog program in \refdef{continuous datalog: ta datalog program},
the entailed fact instance $\langle c, 2 \rangle$ would have the
following set of justifications:
\[
  \{ \ \{ \langle a, 1 \rangle, \langle b, 2 \rangle \},  \{ \langle a, 2 \rangle, \langle b, 1 \rangle \},  \{ \langle a, 2 \rangle, \langle b, 2 \rangle \} \  \}
\]

Having defined the notion of time annotation in Datalog, we now extend
this concept to that of Continuous Datalog (henceforth C-Datalog).

\begin{definition}[Streams]
\labeldef{continuous datalog: stream}
A {\em stream} is a sequence of fact instances ordered
by their instants; in conjunction with the distinction between
multiple instances of a fact by their timestamp, this definition justifies
Axiom~\ref{axiom:continuous datalog: window disjointness}.
\end{definition}

@@

A sliding window of range ${x \in \mathbb{N}}$ over a stream $S$ at
an instant ${i \in \mathbb{T}}$ is the \emph{set} of fact
instances that occur in $S$ during the interval ${(i-x,i]}$, written
${W^{S,x}_{i}}$, and defined as follows:

\[ W^{S,x}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in S \land (i-x) < t \leqslant i \}\]

@@

\begin{axiom}\label{axiom:continuous datalog: window range leq 0}
A window of range $0$ over a stream $S$ at an instant 
${i \in \mathbb{T}}$ is always empty.
\begin{equation*}
W^{S,0}_{i} = \varnothing
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window composition}
A window of range ${x > 0}$ over a stream $S$ at an instant 
${i \in \mathbb{T}}$ is equal to the union of the $x$ consecutive
windows of range $1$ over the stream $S$ preceding and including that at
instant $i$.
\begin{equation*}
W^{S,x}_{i} = \mathop{\cup}_{j=0}^{x-1} W^{S,1}_{i-j}
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window disjointness}
A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an
instant ${i \in \mathbb{T}}$ is disjoint from a second window of 
range ${y \in \mathbb{N}}$ over $S$ at any instant before $i - x$.
\begin{equation*}
W^{S,x}_{i} \cap W^{S,y}_{i-z} = \varnothing
\end{equation*}
where $z \geq x$.
\end{axiom}
